The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-08-06 script/en/chapters/primes.tex
resolved more pdfTeX warnings by adding a chapter specific prefix "xy:" to all cite and bibitem in chapters with bibliography

For the natural numbers, the primes can be compared to elements in chemistry or
the elementary particles in physics (see \cite[p. 22]{Blum1999}).

@@

   This first generalization of the Mersenne
   numbers\index{Mersenne!number!generalized}
   creates the so called Proth numbers\index{Number!Proth}.
   There are (for small $k$) extremely quick prime number tests
   (see \cite{Knuth1981}). This can be performed in practice using software
   such as the Proths software from Yves Gallot\index{Gallot, Yves}:
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://www.prothsearch.net/index.html}
            {\tt http://www.prothsearch.net/index.html}}.
\end{itemize}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{FermatNumbers02}{}
\subsection[Fermat numbers $f(n) = 2^{2^n} + 1$]
              {Fermat numbers\footnotemark~$f(n) = 2^{2^n} + 1$} 
    \footnotetext{%
       The Fermat prime numbers play a role in circle division.
       As proven by Gauss\index{Gauss, Carl Friedrich} a regular $p$-edge
       can only be constructed with the use of a pair of compasses and a 
       ruler, when $p$ is a Fermat prime number.
    }
    \label{L-FermatNumbers02}
    \index{Prime number!Fermat} \index{Fermat!prime number}       
    As mentioned \hyperlink{FermatNumbers01}{above} in 
    chapter~\ref{FermatNumbers01}, Fermat wrote to Mersenne regarding 
    his assumption, that all numbers of this type are primes.
    This assumption was disproved by Euler (1732). The prime $641$ divides $f(5)$\footnote{%
    Surprisingly this number can easily be found by using Fermat's theorem (see e.g. \cite[p. 176]{Scheid1994})
    }. 
%    Surprisingly he would have been able obtain a positive result using
%    the negative prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
f(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
f(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
f(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
f(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
f(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = \mbox{65,537} &  \mapsto {\rm ~prime}  \\
f(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = \mbox{4,294,967,297} = 641 \cdot \mbox{6,700,417} &  \mapsto {\rm ~NOT~prime} ! \\
f(6) = 2^{2^6} + 1  = 2^{64} + 1 &  = \mbox{18,446,744,073,709,551,617} \\
                                 &  = \mbox{274,177} \cdot \mbox{67,280,421,310,721} & \mapsto {\rm ~NOT~prime} !\\
f(7) = 2^{2^7} + 1  = 2^{128} + 1 & = \mbox{(see page~\pageref{F7Morrison})}  
																 & \mapsto {\rm ~NOT~prime} !

@@

Remark: However, what is special about these numbers is that any pair of 
them does not have a common factor other than $1$\footnote{%
This can easily be shown via the following {\em greatest common divisor} ($gcd$) rule $gcd(a,b) = gcd(b-\lfloor b/a \rfloor,a)$ (see page \pageref{NumberTheory_Appendix_A}): %KORR_NumberTheory_Appendix_A mk56
We have for $i<j$: \\
$gcd(e_i,e_j) \le gcd(e_1 \cdots e_i \cdots e_{j-1}, e_j) = gcd(e_j - e_1 \cdots e_i \cdots e_{j-1}, e_1 \cdots e_i \cdots e_{j-1}) 
= gcd(1, e_1 \cdots e_i \cdots e_{j-1}) = 1$.
}. Therefore they are \index{Prime number!relative prime}\index{Relatively prime} {\em relatively prime}.

@@

    There exists no such polynomial that for all $x$ in ${\mathbb Z}$ only
    delivers prime values.
    For a proof of this, please refer to \cite[p. 83 f.]{Padberg1996}, where
    you will also find further details about prime number formulae.

@@

Apart from 2, prime numbers are never even. Apart from 2 and 5, prime numbers
never end in 2, 5 or 0. So you only need to consider numbers ending in 1, 3,
7, 9 anyway (there are infinite primes ending in these numbers; see
\cite[vol. 1, p. 137]{Tietze1973}).

@@

  \item For each number $n$ in ${\bf N}$ there are $n$ consecutive natural
      numbers that are not primes.
      A proof of this can be found in \cite[p. 79]{Padberg1996}.

@@

  \item  \hypertarget{link-Primzahlfunktion-base-a-Proof}{}
      There is a real number a such that the function $f: {\bf N}
      \rightarrow {\mathbb Z}$ where $n \mapsto a^{3^n}$ 
      only delivers primes for all $n$ (see \cite[p. 82]{Padberg1996}).
      Unfortunately, problems arise when we try to determine $a$ 
      (\hyperlink{link-Primzahlfunktion-base-a-Offen}{see below}).

@@

\item Does a polynomial\index{Polynomial} time algorithm exist for 
   calculating the prime factors of a number (see \cite[p. 167]{Klee1997})?
   This question can be divided into the three following questions:
   \begin{itemize} 
      \item Does a polynomial time algorithm exist that decides whether a
            number is prime? \\
	    This question has been answered by the AKS algorithm\index{AKS} 
	    (see chapter \ref{PrimesinP}, ``Primes in P'': 
	    Primality testing is polynomial).

@@

% --------------------------------------------------------------------------
\newpage
\subsection{Orders of magnitude / dimensions in reality}
\label{s:grosord}
In the description of cryptographic protocols and algorithms, numbers 
occur that are so large or so small that they are inaccessible to our 
intuitive understanding. It may therefore be useful to provide comparative 
numbers from the real world around us so that we can develop a feeling for 
the security of cryptographic algorithms. Some of the numbers listed below 
originate from \cite{Schwenk1996} and \cite[p.18]{Schneier1996p}.  
\hypertarget{grosord}{}

@@

\bibitem[Aaronson2003]{Aaronson2003} \index{Aaronson 2003}
    Scott Aaronson, \\
    {\em The Prime Facts: From Euclid to AKS}, \\
    \href{http://www.scottaaronson.com/writings/prime.pdf}
 {\texttt{http://www.scottaaronson.com/writings/prime.pdf}}. \\
    After I had completed this article, I did come across the 
    fine paper by Scott Aaronson, which also offers
    a didactically very well-done introduction to this topic. It is 
    humorous and easy to read but at the same time precise and erudite.

%already defined in elementaryNumberTheory.inc -> 2 davor
\bibitem[Bartholome1996]{2Bartholome1996}  \index{Bartholome 1996}
    A. Bartholom\'e, J. Rung, H. Kern, \\     
    {\em Zahlentheorie f\"ur Einsteiger}, Vieweg 1995, 2nd edition 1996.

\bibitem[Blum1999]{Blum1999} \index{Blum 1999}   
    W. Blum, \\     
    {\em Die Grammatik der Logik}, dtv, 1999.

\bibitem[Bundschuh1998]{Bundschuh1998} \index{Bundschuh 1998}
    Peter Bundschuh, \\
    {\em Einf\"uhrung in die Zahlentheorie}, Springer 1988, 4th edition 1998.

\bibitem[Crandell2001]{Crandell2001} \index{Crandell 2001} \index{Pomerance 2001}
    Richard Crandell, Carl Pomerance, \\
    {\em Prime Numbers. A Computational Perspective}, Springer, 2001.

\bibitem[Doxiadis2000]{Dioxadis2000}
    Apostolos Doxiadis, \\
    {\em Uncle Petros and the Goldbach's Conjecture}, \\
    Faber/Bloomsbury, 2000.

\bibitem[Graham1989]{Graham1989} \index{Graham 1989}     
   R.E. Graham, D.E. Knuth, O. Patashnik, \\
   {\em Concrete Mathematics}, Addison-Wesley, 1989.

\bibitem[Klee1997]{Klee1997} \index{Klee 1997}     
   V. Klee, S. Wagon, \\
   {\em Ungel\"oste Probleme in der Zahlentheorie und der Geometrie der 
   Ebene}, \\ Birkh\"auser Verlag, 1997.

\bibitem[Knuth1981]{Knuth1981} \index{Knuth 1981}     
   Donald E. Knuth, \\ 
   {\em The Art of Computer Programming, vol 2: Seminumerical Algorithms}, \\
   Addison-Wesley, 1969, 2nd edition 1981.

\bibitem[Lorenz1993]{Lorenz1993} \index{Lorenz 1993}     
   F. Lorenz, \\
   {\em Algebraische Zahlentheorie}, BI Wissenschaftsverlag, 1993.

\bibitem[Oppliger2005]{Oppliger2005} \index{Oppliger 2005}
    Rolf Oppliger \\
    {\em Contemporary Cryptography},
    Artech House, 2005, \\
    \href{http://www.esecurity.ch/Books/cryptography.html}
 {\texttt{http://www.esecurity.ch/Books/cryptography.html}}.

\bibitem[Padberg1996]{Padberg1996} \index{Padberg 1996}     
   F. Padberg, \\
   {\em Elementare Zahlentheorie}, 
   Spektrum Akademischer Verlag 1988, 2nd edition 1996.

\bibitem[Pieper1983]{Pieper1983} \index{Pieper 1983}     
   H. Pieper, \\
   {\em Zahlen aus Primzahlen}, 
   Verlag Harri Deutsch 1974, 3rd edition 1983.

\bibitem[Richstein1999]{Richstein1999} \index{Richstein 1999}
    J. Richstein, \\
    {\em Verifying the Goldbach Conjecture up to $4*10^{14},$}
    Mathematics of Computation 70, 2001, p. 1745-1749).

\bibitem[Scheid1994]{Scheid1994} \index{Scheid 1994}
    Harald Scheid, \\ 
    {\em Zahlentheorie}, BI Wissenschaftsverlag, 2nd edition, 1994.

\bibitem[Schneier1996]{Schneier1996p} \index{Schneier 1996}     
    Bruce Schneier, \\
    {\em Applied Cryptography, Protocols, Algorithms, and Source Code in C},\\
    Wiley and Sons, 2nd edition 1996.

\bibitem[Schroeder1999]{Schroeder1999} \index{Schroeder 1999}
    M.R. Schroeder, \\
    {\em Number Theory in Science and Communication}, \\ 
    Springer 1984, 3rd edition 1997, Corrected Printing 1999.

\bibitem[Schwenk1996]{Schwenk1996} \index{Schwenk 1996}     
    J. Schwenk \\
    {\em Conditional Access}, 
    in taschenbuch der telekom praxis 1996, \\
    Hrgb. B. Seiler, Verlag Schiele und Sch\"on, Berlin.

\bibitem[Shoup2005]{Shoup2005} \index{Shoup 2005}
    Victor Shoup \\
    {\em A Computational Introduction to Number Theory and Algebra},\\
    Cambridge University Press, 2005, \\
    \href{http://shoup.net/ntb/}{\texttt{http://shoup.net/ntb/}}.

\bibitem[Tietze1973]{Tietze1973} \index{Tietze 1973}     
    H. Tietze, \\
    {\em Gel\"oste und ungel\"oste mathematische Probleme}, \\
    Verlag C. H. Beck 1959, 6th edition 1973.

>>>>

For the natural numbers, the primes can be compared to elements in chemistry or
the elementary particles in physics (see \cite[p. 22]{pr:Blum1999}).

@@

   This first generalization of the Mersenne
   numbers\index{Mersenne!number!generalized}
   creates the so called Proth numbers\index{Number!Proth}.
   There are (for small $k$) extremely quick prime number tests
   (see \cite{pr:Knuth1981}). This can be performed in practice using software
   such as the Proths software from Yves Gallot\index{Gallot, Yves}:
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://www.prothsearch.net/index.html}
            {\tt http://www.prothsearch.net/index.html}}.
\end{itemize}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{FermatNumbers02}{}
\subsection[Fermat numbers $f(n) = 2^{2^n} + 1$]
              {Fermat numbers\footnotemark~$f(n) = 2^{2^n} + 1$} 
    \footnotetext{%
       The Fermat prime numbers play a role in circle division.
       As proven by Gauss\index{Gauss, Carl Friedrich} a regular $p$-edge
       can only be constructed with the use of a pair of compasses and a 
       ruler, when $p$ is a Fermat prime number.
    }
    \label{L-FermatNumbers02}
    \index{Prime number!Fermat} \index{Fermat!prime number}       
    As mentioned \hyperlink{FermatNumbers01}{above} in 
    chapter~\ref{FermatNumbers01}, Fermat wrote to Mersenne regarding 
    his assumption, that all numbers of this type are primes.
    This assumption was disproved by Euler (1732). The prime $641$ divides $f(5)$\footnote{%
    Surprisingly this number can easily be found by using Fermat's theorem (see e.g. \cite[p. 176]{pr:Scheid1994})
    }. 
%    Surprisingly he would have been able obtain a positive result using
%    the negative prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
f(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
f(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
f(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
f(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
f(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = \mbox{65,537} &  \mapsto {\rm ~prime}  \\
f(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = \mbox{4,294,967,297} = 641 \cdot \mbox{6,700,417} &  \mapsto {\rm ~NOT~prime} ! \\
f(6) = 2^{2^6} + 1  = 2^{64} + 1 &  = \mbox{18,446,744,073,709,551,617} \\
                                 &  = \mbox{274,177} \cdot \mbox{67,280,421,310,721} & \mapsto {\rm ~NOT~prime} !\\
f(7) = 2^{2^7} + 1  = 2^{128} + 1 & = \mbox{(see page~\pageref{F7Morrison})}  
																 & \mapsto {\rm ~NOT~prime} !

@@

Remark: However, what is special about these numbers is that any pair of 
them does not have a common factor other than $1$\footnote{%
This can easily be shown via the following {\em greatest common divisor} ($gcd$) rule $gcd(a,b) = gcd(b-\lfloor b/a \rfloor,a)$ (see page \pageref{NumberTheory_Appendix_GCD}):
We have for $i<j$: \\
$gcd(e_i,e_j) \le gcd(e_1 \cdots e_i \cdots e_{j-1}, e_j) = gcd(e_j - e_1 \cdots e_i \cdots e_{j-1}, e_1 \cdots e_i \cdots e_{j-1}) 
= gcd(1, e_1 \cdots e_i \cdots e_{j-1}) = 1$.
}. Therefore they are \index{Prime number!relative prime}\index{Relatively prime} {\em relatively prime}.

@@

    There exists no such polynomial that for all $x$ in ${\mathbb Z}$ only
    delivers prime values.
    For a proof of this, please refer to \cite[p. 83 f.]{pr:Padberg1996}, where
    you will also find further details about prime number formulae.

@@

Apart from 2, prime numbers are never even. Apart from 2 and 5, prime numbers
never end in 2, 5 or 0. So you only need to consider numbers ending in 1, 3,
7, 9 anyway (there are infinite primes ending in these numbers; see
\cite[vol. 1, p. 137]{pr:Tietze1973}).

@@

  \item For each number $n$ in ${\bf N}$ there are $n$ consecutive natural
      numbers that are not primes.
      A proof of this can be found in \cite[p. 79]{pr:Padberg1996}.

@@

  \item  \hypertarget{link-Primzahlfunktion-base-a-Proof}{}
      There is a real number a such that the function $f: {\bf N}
      \rightarrow {\mathbb Z}$ where $n \mapsto a^{3^n}$ 
      only delivers primes for all $n$ (see \cite[p. 82]{pr:Padberg1996}).
      Unfortunately, problems arise when we try to determine $a$ 
      (\hyperlink{link-Primzahlfunktion-base-a-Offen}{see below}).

@@

\item Does a polynomial\index{Polynomial} time algorithm exist for 
   calculating the prime factors of a number (see \cite[p. 167]{pr:Klee1997})?
   This question can be divided into the three following questions:
   \begin{itemize} 
      \item Does a polynomial time algorithm exist that decides whether a
            number is prime? \\
	    This question has been answered by the AKS algorithm\index{AKS} 
	    (see chapter \ref{PrimesinP}, ``Primes in P'': 
	    Primality testing is polynomial).

@@

% --------------------------------------------------------------------------
\newpage
\subsection{Orders of magnitude / dimensions in reality}
\label{s:grosord}
In the description of cryptographic protocols and algorithms, numbers 
occur that are so large or so small that they are inaccessible to our 
intuitive understanding. It may therefore be useful to provide comparative 
numbers from the real world around us so that we can develop a feeling for 
the security of cryptographic algorithms. Some of the numbers listed below 
originate from \cite{pr:Schwenk1996} and \cite[p.18]{pr:Schneier1996p}.  
\hypertarget{grosord}{}

@@

\bibitem[Aaronson2003]{pr:Aaronson2003} \index{Aaronson 2003}
    Scott Aaronson, \\
    {\em The Prime Facts: From Euclid to AKS}, \\
    \href{http://www.scottaaronson.com/writings/prime.pdf}
 {\texttt{http://www.scottaaronson.com/writings/prime.pdf}}. \\
    After I had completed this article, I did come across the 
    fine paper by Scott Aaronson, which also offers
    a didactically very well-done introduction to this topic. It is 
    humorous and easy to read but at the same time precise and erudite.

%already defined in elementaryNumberTheory.inc -> 2 davor
\bibitem[Bartholome1996]{pr:2Bartholome1996}  \index{Bartholome 1996}
    A. Bartholom\'e, J. Rung, H. Kern, \\     
    {\em Zahlentheorie f\"ur Einsteiger}, Vieweg 1995, 2nd edition 1996.

\bibitem[Blum1999]{pr:Blum1999} \index{Blum 1999}   
    W. Blum, \\     
    {\em Die Grammatik der Logik}, dtv, 1999.

\bibitem[Bundschuh1998]{pr:Bundschuh1998} \index{Bundschuh 1998}
    Peter Bundschuh, \\
    {\em Einf\"uhrung in die Zahlentheorie}, Springer 1988, 4th edition 1998.

\bibitem[Crandell2001]{pr:Crandell2001} \index{Crandell 2001} \index{Pomerance 2001}
    Richard Crandell, Carl Pomerance, \\
    {\em Prime Numbers. A Computational Perspective}, Springer, 2001.

\bibitem[Doxiadis2000]{pr:Dioxadis2000}
    Apostolos Doxiadis, \\
    {\em Uncle Petros and the Goldbach's Conjecture}, \\
    Faber/Bloomsbury, 2000.

\bibitem[Graham1989]{pr:Graham1989} \index{Graham 1989}     
   R.E. Graham, D.E. Knuth, O. Patashnik, \\
   {\em Concrete Mathematics}, Addison-Wesley, 1989.

\bibitem[Klee1997]{pr:Klee1997} \index{Klee 1997}     
   V. Klee, S. Wagon, \\
   {\em Ungel\"oste Probleme in der Zahlentheorie und der Geometrie der 
   Ebene}, \\ Birkh\"auser Verlag, 1997.

\bibitem[Knuth1981]{pr:Knuth1981} \index{Knuth 1981}     
   Donald E. Knuth, \\ 
   {\em The Art of Computer Programming, vol 2: Seminumerical Algorithms}, \\
   Addison-Wesley, 1969, 2nd edition 1981.

\bibitem[Lorenz1993]{pr:Lorenz1993} \index{Lorenz 1993}     
   F. Lorenz, \\
   {\em Algebraische Zahlentheorie}, BI Wissenschaftsverlag, 1993.

\bibitem[Oppliger2005]{pr:Oppliger2005} \index{Oppliger 2005}
    Rolf Oppliger \\
    {\em Contemporary Cryptography},
    Artech House, 2005, \\
    \href{http://www.esecurity.ch/Books/cryptography.html}
 {\texttt{http://www.esecurity.ch/Books/cryptography.html}}.

\bibitem[Padberg1996]{pr:Padberg1996} \index{Padberg 1996}     
   F. Padberg, \\
   {\em Elementare Zahlentheorie}, 
   Spektrum Akademischer Verlag 1988, 2nd edition 1996.

\bibitem[Pieper1983]{pr:Pieper1983} \index{Pieper 1983}     
   H. Pieper, \\
   {\em Zahlen aus Primzahlen}, 
   Verlag Harri Deutsch 1974, 3rd edition 1983.

\bibitem[Richstein1999]{pr:Richstein1999} \index{Richstein 1999}
    J. Richstein, \\
    {\em Verifying the Goldbach Conjecture up to $4*10^{14},$}
    Mathematics of Computation 70, 2001, p. 1745-1749).

\bibitem[Scheid1994]{pr:Scheid1994} \index{Scheid 1994}
    Harald Scheid, \\ 
    {\em Zahlentheorie}, BI Wissenschaftsverlag, 2nd edition, 1994.

\bibitem[Schneier1996]{pr:Schneier1996p} \index{Schneier 1996}     
    Bruce Schneier, \\
    {\em Applied Cryptography, Protocols, Algorithms, and Source Code in C},\\
    Wiley and Sons, 2nd edition 1996.

\bibitem[Schroeder1999]{pr:Schroeder1999} \index{Schroeder 1999}
    M.R. Schroeder, \\
    {\em Number Theory in Science and Communication}, \\ 
    Springer 1984, 3rd edition 1997, Corrected Printing 1999.

\bibitem[Schwenk1996]{pr:Schwenk1996} \index{Schwenk 1996}     
    J. Schwenk \\
    {\em Conditional Access}, 
    in taschenbuch der telekom praxis 1996, \\
    Hrgb. B. Seiler, Verlag Schiele und Sch\"on, Berlin.

\bibitem[Shoup2005]{pr:Shoup2005} \index{Shoup 2005}
    Victor Shoup \\
    {\em A Computational Introduction to Number Theory and Algebra},\\
    Cambridge University Press, 2005, \\
    \href{http://shoup.net/ntb/}{\texttt{http://shoup.net/ntb/}}.

\bibitem[Tietze1973]{pr:Tietze1973} \index{Tietze 1973}     
    H. Tietze, \\
    {\em Gel\"oste und ungel\"oste mathematische Probleme}, \\
    Verlag C. H. Beck 1959, 6th edition 1973.
